Contracts for unions and intersections are hard for Nickel

### 2021-05-17 paper.tex
[WIP] Start filling the optimizations section

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
historical bit should probably be moved to related works eventually} Nickel
adopts a similar type system, with both statically typed terms, dynamically
typed terms, and first-class contracts. In "first-class contracts", we include
the ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability to write a small
validation function for, say, a port number, to have it immediately available as
a contract that can be composed with other type constructors\unsure{Needs code
to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
system.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:union-encoding}
\medskip
\small
Defines a port contract from a predicate and use it in the definition of a
\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
contract to a value, similar to `:` for static types. We assume that a
\lstinline+Host+ contract has also been defined.
\end{figure}

In this hybrid system, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that checks at run-time that
a value is indeed a member of the corresponding type. Contracts for various
extensions of the simply typed $\lambda$-calculus have been investigated
\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
Amongst all the common extensions to the polymorphic lambda calculus, one of the
most useful and widespread turns out to be one of the most challenging:
\emph{union types}.

>>>>

The good performances of modern programs heavily rely on complex optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, because the functional which model is very different from what actual
computers are good at executing, naive code generation can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and function
is a costly operation from the point of view of low-level execution.

One such crucial optimization is inlining. Inlining eliminates a function
application by substituting the function for its definition at compile time (or
before execution, for interpreted language). This is especially important for
short functions that are called repeatedly.

% EXAMPLE INLINING

Another optimization is common subexpression elimination (CSE). CSE somehow
goes into the other direction: it factors out equal subexpressions to avoid
recomputing them several times.

% EXAMPLE CSE

Beyond CSE, there exists other similar transformations, such as Haskell's
let-floating~\ref{} or loop-invariant code motion, which amount to hoist
invariant out from repeated blocks.

More generally, there is a whole family of opitmizations, such as let-floating,
\begin{lstlisting}[language=caml]
type Ordering = Smaller | Equals | Greater
sort : 'a list
       -> ('a -> 'a -> Ordering)
       -> 'a list
random_list : unit -> int list

sort (random_list ()) (fun x y ->
  if x < y then Smaller
  else if x == y then Equals
  else Greater)
\end{lstlisting}

Common subexpression elemination (CSE) etc.

\begin{lstlisting}[language=caml]

\end{lstlisting}

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
historical bit should probably be moved to related works eventually} Nickel
adopts a similar type system, with both statically typed terms, dynamically
typed terms, and first-class contracts. In "first-class contracts", we include
the ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability to write a small
validation function for, say, a port number, to have it immediately available as
a contract that can be composed with other type constructors\resolved{Needs code
to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
system. An example is given in Figure~\ref{fig:contract-composition}. The first
line defines a port contract from a predicate and use it in the definition of a
larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
entails that a value respect a contract, akin to \lstinline+:+ for types. We
assume that a \lstinline+Host+ contract has been defined in a similar way.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\medskip
\small
Defines a port contract from a predicate and use it in the definition of a
\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
contract to a value, similar to `:` for static types. We assume that a
\lstinline+Host+ contract has also been defined.
\end{figure}

In this hybrid system, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that checks at run-time that
a value is a member of the corresponding type. Contracts for various
extensions of the simply typed $\lambda$-calculus have been investigated
\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
Amongst all the common extensions to the polymorphic lambda calculus, one of the
most useful and widespread turns out to be one of the most challenging:
\emph{union types}.
